---
title: Freezing Python versions inside Bioconductor packages
author: 
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: 5 October 2019"
output:
  BiocStyle::html_document:
    toc_float: true
package: basilisk
bibliography: 
vignette: >
  %\VignetteIndexEntry{Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(basilisk)
```

# Why?

Packages like `r CRANpkg("reticulate")` make it easy to use Python modules in our R-based data analyses, allowing us to leverage Python's strengths in fields such as machine learning. 
However, it is notoriously difficult to ensure that a consistent version of Python is available with a consistently versioned set of modules, especially when the system installation of Python is used.
As a result, we cannot easily guarantee that some Python code executed via `r CRANpkg("reticulate")` on one computer will yield the same results as the same code run on another computer.
It is also possible that two R packages depend on incompatible versions of Python modules, such that it is impossible to use both packages at the same time.
These versioning issues represent a major inhibitor on the ability of R/Bioconductor packages to reliably execute Python code across a variety of systems.

# What?

`r Biocpkg("basilisk")` provides a self-contained Python installation that is fully managed by the Bioconductor installation machinery.
This allows developers of downstream Bioconductor packages to be sure that they are working with the same version of Python on all systems.
Additionally, `r Biocpkg("basilisk")` will only install modules from a cohort of pre-defined versions, thus ensuring that the same modules are used by all downstream packages.
(This cohort is currently defined as the latest version of each package up to the 5th October, 2019.)
Consistency in the execution environment enables reproducible analysis, simplifies debugging of code and improves interoperability between compliant packages.

# How?

Consider a downstream Bioconductor package _X_ that depends on, say, `r PyPiLink("scikit-learn")`.
Any function in _X_ that depends on the installed modules should call:

```{r}
useBiocPython()
X <- importModule("sklearn.ensemble")
names(X)
```

The `useBiocPython()` call ensures that the Bioconductor-managed instance of Python is used rather than the system installation.
The `importModule()` call ensures that `r PyPiLink("scikit-learn")` is installed before any import is attempted.
We can then use standard `r CRANpkg("reticulate")` code to call specific functions in the loaded module.

`r Biocpkg("basilisk")` is primarily intended for developers of R/Bioconductor packages, **not** for end-users.
Nonetheless, users can call these functions directly to add modules to the local Python instance.

# Session information

```{r}
sessionInfo()
```
